---
title: "Effect of low sample size on genetic results"
author: "Luke Warkentin"
date: "2023-03-30"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
options(scipen = 1000000)
```

### Make some fake data

```{r make data}

pop_a <- 300 # Number of fish in population a
pop_b <- 50000 # number of fish not in population a
n_catch <- sum(pop_a, pop_b) # total fish caught
fish_caught <- c(rep(1, pop_a), rep(0, pop_b)) # vector of 1 and 0, 1 if fish is population a
true_prop <- pop_a / n_catch  # 'true' proportion of population a fish
min_sample_size <- 10
max_sample_size <- 5000
step_sample_size <- 30
n_reps_sample <- 500
sample_size <- rep(seq( min_sample_size , max_sample_size, by=step_sample_size), each= n_reps_sample)
hypo_sample_rate <- 0.005  # hypothetical low sample rate to compare on figures, as a reference

# Get estimated proportion population a by sampling fish caught with repeated samples of various sample sizes
estimated_prop <- sapply( sample_size, function(x) {
                          sum(sample(fish_caught, size = x)) / x
                          }
        )

s <- data.frame(sample_size, estimated_prop)
s$estimated_pop_a <- s$estimated_prop * n_catch


s_sum <- s %>% group_by(sample_size) %>% summarise(mean_pop_a = mean(estimated_pop_a, na.rm=TRUE), 
                                                 median_pop_a = median(estimated_pop_a, na.rm=TRUE),
                                                 q_up = quantile(estimated_pop_a, 0.05), 
                                                 q_low = quantile(estimated_pop_a, 0.95))

```

``` {r setup for graphs, include=FALSE}
quantile_colour <- "springgreen"
median_colour <- "springgreen"
mean_colour <- "red3"

```


Simulated data:

- Total catch is `r n_catch` Chinook in fishery
- There are `r pop_a` Chinook from population **a** (low encounter rate) in the total catch. 
- For illustration purposes, compare a hypothetical sample rate of `r hypo_sample_rate` or `r hypo_sample_rate * 100`\% with simulated data

Simulated sampling is done `r n_reps_sample` each times for sampling rates between `r min_sample_size` and `r max_sample_size`. 

Assumptions: 

- Fish are individually identifiable to population with 100% accuracy.
- Sampling is completely random.

### Figures

``` {r plot full y axis, echo=FALSE, fig.cap="Effect of sample size on estimated size of population a. Blue line is the 'true' population a size.  Red bars are mean population. Green line and shaded area are median with 5% and 95% percentiles."}

ggplot() +
  #geom_violin(data = s,aes(x = sample_size, y = estimated_pop_a, group = sample_size), fill="gray", colour="gray")   +
  geom_point( data = s,aes( x = sample_size, y = estimated_pop_a), shape=1, size=2, position = position_jitter(width=3)) +
  geom_vline(aes(xintercept = hypo_sample_rate * n_catch), colour="gray", linetype=2) +
  geom_text(aes(x=hypo_sample_rate * n_catch + 5, y=5000, label=paste0("Sample rate = ", hypo_sample_rate*100, "%")), hjust = 0, colour="gray") +
  geom_ribbon(data = s_sum , aes(ymin=q_up, ymax=q_low, x=sample_size), alpha=0.5, fill=quantile_colour) +
    geom_line(data = s_sum , aes(y=median_pop_a, x=sample_size), linewidth=1.2, colour=median_colour) +
geom_point(data = s_sum , aes(y=mean_pop_a, x=sample_size), colour=mean_colour, shape=45, size=4, stroke=3) +
  ylab("Estimated fish in population a") + 
  xlab("Sample size") +
  geom_hline(aes(yintercept=0)) +
  geom_hline(aes(yintercept=pop_a), colour="dodgerblue") +
  #geom_text(aes(x=500, y=pop_a-200, label=paste0("Actual population = ", pop_a)),vjust=0, colour="dodgerblue") + 
  coord_cartesian() +
  theme_classic()

```

\linebreak


``` {r plot truncated, echo=FALSE, fig.cap="Effect of sample size on estimated size of population a, x and y axes truncated. Blue line is the 'true' population a size. Red bars are mean population. Green line and shaded area are median with 5% and 95% percentiles." }

ggplot() + 
  geom_violin(data = s,aes(x = sample_size, y = estimated_pop_a, group = sample_size), fill="gray", colour="gray")   +
  geom_ribbon(data = s_sum , aes(ymin=q_up, ymax=q_low, x=sample_size), alpha=0.3, fill=quantile_colour) +
    geom_line(data = s_sum , aes(y=median_pop_a, x=sample_size), linewidth=1.2, colour=median_colour) +
geom_point(data=s, aes( x = sample_size, y = estimated_pop_a), shape=1, size=2, position = position_jitter(width=3)) +
  geom_vline(aes(xintercept = hypo_sample_rate * n_catch), colour="gray", linetype=2) +
  geom_text(aes(x=hypo_sample_rate * n_catch + 5, y=2000, label=paste0("Sample rate = ", hypo_sample_rate*100, "%")), hjust = 0, colour="gray") +
  geom_point(data = s_sum , aes(y=mean_pop_a, x=sample_size), colour=mean_colour, shape=45, size=6, stroke=6) +
 geom_hline(aes(yintercept=0)) +
  ylab("Estimated fish in population a") +
  xlab("Sample size") +
  geom_hline(aes(yintercept=pop_a), colour="dodgerblue") +
  #geom_text(aes(x=200, y=pop_a-300, label=paste0("Actual population = ", pop_a)), vjust=0, colour="dodgerblue") + 
  coord_cartesian(xlim=c(0,1000), ylim = c(-20,2000)) +
  theme_classic()

```

\linebreak

 ``` {r plot log 10 y axis, echo=FALSE, warning=FALSE, fig.cap= "Effect of sample size on estimated size of population a, y axis is on log 10 scale. Blue line is the 'true' population a size.  Red bars are mean and purple points are median estimated population, by sample size."   }
# 
# max_est <- max(s$estimated_pop_a)
# 
# ggplot(data = s, aes( x = sample_size, y = estimated_pop_a)) +
#   #geom_violin(aes(group = sample_size), fill="gray", colour="gray")   +
#   geom_point( aes( x = sample_size, y = estimated_pop_a), shape=1, size=2, position = position_jitter(width=3)) +
#   geom_vline(aes(xintercept = hypo_sample_rate * n_catch), colour="gray", linetype=2) +
#   geom_text(aes(x=hypo_sample_rate * n_catch + 5, y=50000, label=paste0("Sample rate = ", hypo_sample_rate*100, "%")), hjust = 0, colour="gray") +  
#   geom_point(data = s_sum , aes(y=mean_pop_a, x=sample_size), colour="gray") +
#   geom_point(data = s_sum , aes(y=median_pop_a, x=sample_size), colour=quantile_colour) +
#   ylab("Estimated fish in population a") +
#   geom_hline(aes(yintercept=0)) +
#   scale_y_log10() +
#   geom_hline(aes(yintercept=pop_a), colour="dodgerblue") +
#   #geom_text(aes(x=200, y=pop_a, label=paste0("Actual population = ", pop_a)), vjust=0, colour="dodgerblue") + 
#   coord_cartesian(xlim=c(0,1000)) +
#   theme_classic()
# 
 ```



``` {r plot log 10 y axis full x, echo=FALSE, warning=FALSE, fig.cap="Effect of sample size on estimated size of population a, y axis is on log 10 scale, x axis not truncated. Blue line is the 'true' population a size. Red bars are mean population. Green line and shaded area are median and 5% and 95% percentiles."  }
max_est <- max(s$estimated_pop_a)

ggplot() +
   geom_point(data=s, aes( x = sample_size, y = estimated_pop_a), shape=1, size=2, position = position_jitter(width=3)) +
  geom_vline(aes(xintercept = hypo_sample_rate * n_catch), colour="gray", linetype=2) +
  geom_ribbon(data = s_sum , aes(ymin=q_up, ymax=q_low, x=sample_size), alpha=0.3, fill=quantile_colour) +
   geom_line(data = s_sum , aes(y=median_pop_a, x=sample_size), linewidth=1.2, colour=median_colour) +
geom_text(aes(x=hypo_sample_rate * n_catch + 5, y=50000, label=paste0("Sample rate = ", hypo_sample_rate*100, "%")), hjust = 0, colour="gray") +
  geom_hline(aes(yintercept=0)) +    
  geom_point(data = s_sum , aes(y=mean_pop_a, x=sample_size), colour=mean_colour, shape=45, size=4, stroke=3) +
  ylab("Estimated fish in population a") +
  xlab("Sample size") +
 geom_hline(aes(yintercept=0)) +
  scale_y_log10()+
  geom_hline(aes(yintercept=pop_a), colour="dodgerblue") +
  #geom_text(aes(x=200, y=pop_a, label=paste0("Actual population = ", pop_a)), vjust=0, colour="dodgerblue") + 
  coord_cartesian() +
  theme_classic()

```
